Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-10 Ben_Manuscripts/hdphmm/Draft.tex
Updates to methods and clustering

  % MRS: going to JPC, so emphasize the physical systems.
  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams.
  \begin{itemize}
    \item Desalination and boric acid removal from seawater.
    \item Organic micropollutants
    \item Hydraulic fracturing flowback water
    \item Breathable chemical warfare agent barriers
    \item High purity chemicals
    \item While many researchers focus on membrane permeability, we may be 
    able to reduce costs of commercial nanofiltration and reverse osmosis with
    higher selectivity.~\cite{werber_materials_2016}
  \end{itemize}

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables.
  %MRS: should explicitly point out that just calculating diffusion coefficients don't work for subdiffusive behavior, so
  %MRS need a different way to do it.
  %BJC2: added last two bullet points
  \begin{itemize}
  	\item If one could apply molecular dynamics (MD) simulations to explain important
  	experimental metrics such as solute flux and selectivity in terms of detailed 
  	chemically dependent solute motion, then we can more easily bridge the disconnect
  	between theory and application.
  	\item In many systems, it's possible to achieve this by calculating diffusion 
  	constants.\cite{}
  	\item Unfortunately, for systems which exhibit complex subdiffusive dynamics on the
  	time scales accessible to MD simulation, one cannot calculate reliable diffusion 
  	constants since solute mean square displacements (MSDs) are non-linear 
        %MRS2: added
        on timescales accessible to molecular simulation.
  	\item Therefore, we are in need of a more flexible approach for modeling solute 
  	dynamics which makes less assumptions about the solute's average behavior.
  	%BJC2: I'm moving these towards the end
%  	\item There are two main challenges associated with this goal.
%  	%since complex and
%  	%diverse diffusive motion on time scales accessible to molecular simulations prevents
%  	%us from easily relating chemically dependent solute motion to experimental time scales.
%  	\item First, we need a simple way to identify and parameterize the fluctuations
%  	characteristic to dominant solute-membrane interactions.
%  	\item Second, we need a way to propagate this behavior far beyond the time scales
%  	afforded by MD.
  \end{itemize}

  %MRS: point out that LLC's are examples where there are lots of hopping/jumping (cite previous papers) and
  %MRS: so we need different approaches.
  %BJC2: added last bullet point to lead in to following paragraph
  In this work, we are particularly interested in the design of lyotropic liquid crystal (LLC)
  monomers, a class of amphiphilic molecules whose ordered self-assembled phases can be 
  cross-linked into mechanically strong membranes capable of highly selective separations.
  \begin{itemize}
  	\item Inverted hexagonal, H\textsubscript{II}, phase LLC membranes are characterized
  	by hexagonally packed, uniform-sized and straight pores, an ideal geometry for high 
  	throughput transport.
	\item Their pores are lined with the LLC monomer functional groups which can
	potentially be designed to interact with solutes in a chemically-specific manner.
	\item It would be extremely useful to experimentalists if we could connect LLC monomer
	design with measurable quantities.
	\item This goal is complicated by the non-Brownian hopping and trapping behavior
	induced by the membrane's diverse topology and inhomogeneous structure.~\cite{coscia_understanding_2019,coscia_chemically_2019}
  \end{itemize}

@@

  In this work we attempt to minimize human effort by applying the infinite hidden 
  Markov Model (IHMM)~\cite{beal_infinite_2002,calderon_inferring_2015}, a nonparameteric 
  Bayesian classification technique, in order to automatically detect and infer the 
  parameters of an unknown number of latent vector autoregressive (VAR) modes present 
  in solute center-of-mass time series trajectories. 
  \begin{itemize}
    %MRS: start with where it has been used before (to establish utility) a
    \item It's been used extensively for speaker diarization and has found some application
    in single particle trajectory analysis.
    %MRS: Then say it hasn't been used in this context (do we know of any, then cite, if none we have found, say so.)
    \item The IHMM has not been used much for MD trajectory analysis.

@@

  We aim to address two primary scientific 
  questions with this work. 
  First, we want
  to know if the IHMM algorithm can help researchers efficiently uncover underlying
  transport mechanisms which give rise to different solute dynamical behavior.
  Second, we want to predict macroscopic transport by using our model to generate 
  stochastic solute trajectories at much longer timescales than possible with MD simulations, 
  extending into timescale where diffusive behavior again occurs.

  We use the parameters of the states identified by the IHMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms. We cluster similar 
  state parameters in order to reduce the state space and to understand which segments 
  of the solute trajectories exhibit similar behavior. We support our mechanistic 
  hypotheses via quantitative comparison to various physical membrane properties and
  solute-membrane interactions.

  Finally, we use the VAR parameters and state transition probability matrix in
  order to generate stochastic trajectory realizations with behavior similar to that
  exhibited by MD. 
  \begin{itemize}  
    \item One can project these realizations onto much longer timescales with 
    computational ease and use them to predict macroscopic properties.
  \end{itemize}

@@

  Hidden Markov models (HMMs) are a useful and widely used technique for modeling
  sequences of observations where the probability of the next observation in a 
  sequence depends, at least in part, on a previous unobserved, latent or hidden,
  state.~\cite{beal_infinite_2002} In the context of our simulations, the observations
  correspond to the center of mass coordinates of the solutes versus time, and the
  states correspond to 
  % MRS2: added - ``states'' aren't quite ``behavior''
  segments of the trajectory with similar
  dynamical behavior which gives rise to those types
  of observations. The probability of transitioning to a state based on the current
  state is 
  %MRS2: redundant
  % mathematically 
  defined in terms of an $n\times n$ transition probability
  matrix, $T$, where $n$ is the number of states. Unfortunately, standard HMMs 
  require $n$ to be known \textit{a priori}. One can partially overcome this by 
  testing a range of numbers of hidden states and determining which is the best 
  representation of the data.~\cite{pohle_selecting_2017}

  %BJC: I'm not sure how much of this I should include. This is pretty much 
  % a reproduction of Fox et al's explanation: https://ieeexplore.ieee.org/abstract/document/5563110
  %MRS2: this is fine.
  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.~\cite{fox_bayesian_2010} Using some 
  base probability distribution, $H$, a Dirichlet process (DP) generates discrete 
  distributions, $G_0$, over a countably infinite number of probability measures:
  \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
  \end{equation}
  where the $\theta_k$ are values drawn from the base distribution and the weights
  $\beta_k$ come from a stick-breaking process parameterized by the concentration 
  parameter $\gamma$ (equivalently referred to as GEM($\gamma$)).~\cite{halmos_random_1944} In shorthand, this
  specification can be expressed as $G_0 \sim DP(\gamma, H)$. The concentration 
  parameter, $\gamma$, expresses one's confidence in the base distribution $H$.
  We use a uniform base distribution for $H$. When $\gamma\to 0$, the first 
  weight of $G_0$, $\beta_1$, approaches unity and for $\gamma\to\infty$, the weights
  become uniform and $G_0$ closely resembles $H$. Each row, $G_j$, of the transition 
  matrix is produced by drawing from a DP specified using the $\beta$ vector as a 
  discrete base distribution and a separate concentration parameter, $\alpha$.
  \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
  \end{equation}
  This hierarchical specification ensures that the transition probabilities in 
  each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
  Once the model has converged only a finite number of states will have significant
  sampling.

@@

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a Python implementation of this process which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  %MRS2: give location for the code here where you introduce it.
  Therefore, we looked for convergence of the parameters in Equation~\ref{eqn:var}.
  % each entry in the $\mathbf{c}$ vector as well as the $A$ and $\Sigma$ matrices of Equation~\ref{eqn:var}. 
  We detected equilibration 
  of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

  %BJC2: I say 'applied' a lot. Diversity that eventually
  We carefully applied the IHMM algorithm in a way which takes advantage of the
  system's cylindrical symmetry. While there
  are a number of viable ways in which one could choose to analyze these time series,
  we chose to use a multi-step procedure that we believe adequately distinguishes 
  every type of distinct dynamical behavior exhibited by the solutes. The final
  parameterization of the model is in cylindrical coordinates since each pore is a
  cylinder where we expect solutes to exhibit radially symmetric dynamics. We 
  present a step-by-step demonstration of the procedure that follows in 
  Section~\ref{S-section:ihmm_procedure} of the Supporting Information.

  We start by applying the IHMM algorithm to the 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center. We tracked the solute's
  motion along the pore axis with the center-of-mass $z$ coordinate. Using the nearest
  pore center as the origin, we represented the radial distance of each solute's 
  center-of-mass from the pore center in 2 dimensions, $x$ and $y$. By working in 
  Cartesian coordinates, we avoid mathematical complexity introduced by cylindrical 
  coordinates while estimating the state sequence.

  We applied the IHMM to each of the 24 solute trajectories independently.
  Although the IHMM is capable of identifying an infinite number of states, 
  a Dirichlet process tends to exhibit a ``rich get richer" effect, favoring
  a fewer number of states.~\cite{dreyer_discovering_2011} By applying the algorithm to each trajectory 
  independently, we reduce the possibility of lumping together multiple 
  similar states which we would prefer to stay separated before clustering.
  Note that the states identified by the IHMM are heavily influenced by 
  the Gaussian prior placed on $\mathbf{c}$ in Equation~\ref{eqn:var}. 
  In Section~\ref{S-section:prior_guesses} of the Supporting Information we 
  outline our method for choosing reliable prior parameters.

@@

  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) due to the system's radial 
  symmetry. We converted the $x$ and $y$ center-of-mass coordinates to $r$. 
  We fixed the state sequence from the 3D parameterization and applied the 
  inference component of the IHMM procedure to the cylindrical trajectories 
  in order to estimate the VAR(1) parameters in terms of $r$ and $z$. Since
  we fixed the state sequence, and we are using conjugate priors, the VAR 
  parameters converge very quickly (see Figure~\ref{S-fig:fixed_state_convergence}).
  Therefore, we only ran the inference procedure for 100 iterations. 
  % MRS: hmm. Here the new covariance parameters is redone in r - does there need to be any transformation applied? 
  % MRS: though might be OK, since x^2 + y^2 = r^2 so maybe covariance are additive if x and y uncorrelated (which they should be)
  % MRS2: As I mentioned before, the issue is if the likelihood for the covariance matrix is properly defined for radial diffusion.  We can check over the equations.  It likely wouldn't change the qualitative results much, but should ne checked.  Would not affect the clustering.
  We defined the final parameters of each state as the mean of the parameters 
  from each iteration recorded after the equilibration time point. We found the 
  equilibration time point of each component of the $\mathbf{c}$ vector as well as 
  $A$ and $\Sigma$ matrices, and then used the longest equilibration time of all 
  dimensions as the equilibration time point.

  \subsection{Generating Stochastic IHMM Trajectory Realizations}\label{method:realizations}

  %BJC: I'll need to expand on this a bit to make it clear. Talk about the bootstrapping etc.
  Finally, we generated stochastic trajectory realizations using the finalized
  parameter sets. We drew state sequences with transition probabilities given by
  $T$. While in a given state, we simulated motion according to the VAR(1)
  parameterization of that state. After each state transition, we reset the 
  unconditional mean of each state based on the particle's position immediately
  before the state transition occurred.

@@

  %MRS2: maybe add subsubsections naming/labeling each part of the process so that people have a sense of the steps,
  %MRS2: makes it easier to read.
  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size. For each solute studied, we identified 200-325
  independent states, each with separate VAR(1) parameters. Many of these states
  exhibit very similar dynamical behavior except their mean levels in $r$ and $z$
  are different.

  We reduced the parameter space used agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage 
  criteria, which works to minimize the sum of the squared differences within all
  clusters.~\cite{ward_hierarchical_1963} We elected to choose the number of clusters
  rather than the distance threshold. For our data, non-parametric methods such as 
  Bayesian Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013}
  tend to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information).
  %MRS: not clear what the consequence of the above is?

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the radial means, $\mu_r$, 
  and the self-transition probabilities, $T_{ii}$, of each state for a total of 6 clustering
  dimensions. The means in $r$ are likely meaningful since the membrane is 
  %MRS2: added
  radially inhomogeneous from each pore, 
  transitioning from hydrophilic pores to hydrophobic tails. 
  %MRS2: 
  However, the channels are isotropic in $z$ so we do not cluster on $\mu_z$.
  To cluster on $T_{ii}$, we first
  cast it in terms of the expected dwell times. The expected value of the number of sequential
  self-transitions is simply $\frac{1}{1 - T_{ii}}$. This relationship correctly implies that 
  dwell times approach infinity as $T_{ii}$ approaches 1. There we found the most success with
  agglomerative clustering by linearizing this relationship and clustering on $-\log(1 - T_{ii})$
  (see Section\ref{S-section:selfT_cluster} of the Supporting Information for a graphical
  illustration).

@@

%  Based on these guidelines, we decided to 
%  group the $A$ and $\Sigma$ parameters into five clusters each and the radial
%  means into three clusters. Of the 75 distinct states allowed by this formulation,
%  the solutes in this study showed behavior from 32--36 distinct clusters. While
%  this may seem like a high number of states, many are sampled infrequently.
  %MRS: what is the consequence of being sampled infrequently.

@@

  We remapped the state sequence based on the cluster assignments and parameterized
  the clusters using the inference component of the IHMM algorithm. 
  First, we modified the ($r$, $z$) solute trajectories so that they had a mean of zero,
  leaving only the fluctuations. We did this by subtracting the mean from each 
  same-state segment of the unclustered trajectory. We used the IHMM algorithm on this
  modified trajectory to infer the clustered state parameters by fixing the clustered 
  state sequence.

  We obtained $\mathbf{c}$ vectors of the clustered states by averaging 
  each value of $\mathbf{c}$ assigned to the same cluster. Note that we only care
  about the $r$ component of $\mathbf{c}$ because solute trajectories are not 
  bound in the $z$ direction.

@@

  %We developed a way to 
  We measure the local density of the membrane at arbitrary
  points in the unit cell as follows. We histogrammed the positions in three dimensions
  Since our system is in a monoclinic unit cell, we periodically replicated
  the system in the $\pm$$x$, $y$ and $z$ directions and then chose the bounds on the
  histogram in order to create a rectangular box encompassing the unit cell with
  a 1 nm buffer between the histogram and unit cell boundaries.~\cite{van_der_walt_numpy_2011} We then used a regular
  grid interpolator in order to allow interpolation at arbitrary points within the grid.~\cite{virtanen_scipy_2020}

@@

  \begin{figure}
  \centering
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_MET.pdf}
  \caption{methanol}\label{fig:unclustered_msd_MET}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_GCL.pdf}
  \caption{ethylene glycol}\label{fig:unclustered_msd_GCL}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_URE.pdf}
  \caption{urea}\label{fig:unclustered_msd_URE}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_ACH.pdf}
  \caption{acetic acid}\label{fig:unclustered_msd_ACH}
  \end{subfigure}
  \caption{The MD MSDs (black) of methanol and urea are under-predicted by 
  the IHMM (orange), while the predictions match within uncertainty for ethylene
  glycol and urea.}\label{fig:unclustered_msds}
  %MRS2: to save space, could consider the graphs in 1x4 across the top, with a single legend (since all have the same), and slightly thicker lines  %There are only 2 lines per plot, so it's not too noisy to be that size
  \end{figure}

>>>>

  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams.
  \begin{itemize}
    \item Desalination and boric acid removal from seawater.
    \item Organic micropollutants
    \item Hydraulic fracturing flowback water
    \item Breathable chemical warfare agent barriers
    \item High purity chemicals
    \item While many researchers focus on membrane permeability, we may be 
    able to reduce costs of commercial nanofiltration and reverse osmosis with
    higher selectivity.~\cite{werber_materials_2016}
  \end{itemize}

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables.
  \begin{itemize}
  	\item If one could apply molecular dynamics (MD) simulations to explain important
  	experimental metrics such as solute flux and selectivity in terms of detailed 
  	chemically dependent solute motion, then we can more easily bridge the disconnect
  	between theory and application.
  	\item In many systems, it's possible to achieve this by calculating diffusion 
  	constants.%\cite{}
  	\item Unfortunately, for systems which exhibit complex short timescale 
  	subdiffusive dynamics, one cannot calculate reliable diffusion 
  	constants since solute mean square displacements (MSDs) are non-linear 
    on the timescales accessible to molecular simulation.
  	\item Therefore, we are in need of a more flexible approach for modeling solute 
  	dynamics which makes less assumptions about the solute's average behavior.
  \end{itemize}

  In this work, we are particularly interested in the design of lyotropic liquid crystal (LLC)
  monomers, a class of amphiphilic molecules whose ordered self-assembled phases can be 
  cross-linked into mechanically strong membranes capable of highly selective separations.
  \begin{itemize}
  	\item Inverted hexagonal, H\textsubscript{II}, phase LLC membranes are characterized
  	by hexagonally packed, uniform-sized and straight pores, an ideal geometry for high 
  	throughput transport.
	\item Their pores are lined with the LLC monomer functional groups which can
	potentially be designed to interact with solutes in a chemically-specific manner.
	\item It would be extremely useful to experimentalists if we could connect LLC monomer
	design with measurable quantities.
	\item This goal is complicated by the non-Brownian hopping and trapping behavior induced
	by the membrane's diverse topology and inhomogeneous structure.~\cite{coscia_understanding_2019,coscia_chemically_2019}
  \end{itemize}

@@

  In this work we attempt to minimize human effort by applying the infinite hidden 
  Markov Model (IHMM)~\cite{beal_infinite_2002,calderon_inferring_2015}, a nonparameteric 
  Bayesian classification technique, in order to automatically detect and infer the 
  parameters of an unknown number of latent vector autoregressive (VAR) modes present 
  in solute center-of-mass time series trajectories. 
  \begin{itemize}
    %MRS: start with where it has been used before (to establish utility) a
    \item It's been used extensively for speaker diarization and has found some application
    in single particle trajectory analysis.~\cite{calderon_inferring_2015,fox_sticky_2011}
    %MRS: Then say it hasn't been used in this context (do we know of any, then cite, if none we have found, say so.)
    \item The IHMM has not been used much for MD trajectory analysis.

@@

  We aim to address two primary scientific questions with this work:
  First, we want to know whether it is possible to predict macroscopic transport
  properties. We use the VAR parameters and state transition probability matrix of
  the IHMM in order to generate stochastic trajectory realizations with behavior 
  similar to that exhibited by MD. One can project these realizations onto much 
  longer timescales with computational ease extending into timescales where 
  regular diffusive behavior again occurs.
  Second, we want to know if the IHMM algorithm can help researchers
  efficiently uncover underlying transport mechanisms which give rise to different 
  solute dynamical behavior.
  We use the parameters of the states identified by the IHMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms. We cluster similar 
  state parameters in order to reduce the state space and to understand which segments 
  of the solute trajectories exhibit similar behavior. We support our mechanistic 
  hypotheses via quantitative comparison to various physical membrane properties and
  solute-membrane interactions.

@@

  Hidden Markov models (HMMs) are a useful and widely used technique for modeling
  sequences of observations where the probability of the next observation in a 
  sequence depends, at least in part, on a previous unobserved, latent or hidden,
  state.~\cite{beal_infinite_2002} In the context of our simulations, the observations
  correspond to the center of mass coordinates of the solutes versus time, and the
  states correspond to segments of the trajectory with similar dynamical behavior 
  which gives rise to those types of observations. The probability of transitioning 
  to a state based on the current state is defined in terms of an $n\times n$ 
  transition probability matrix, $T$, where $n$ is the number of states. Unfortunately,
  standard HMMs require $n$ to be known \textit{a priori}. One can partially overcome
  this by testing a range of numbers of hidden states and determining which is the
  best representation of the data.~\cite{pohle_selecting_2017}

  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.~\cite{fox_bayesian_2010} Using some 
  base probability distribution, $H$, a Dirichlet process (DP) generates discrete 
  distributions, $G_0$, over a countably infinite number of probability measures:
  \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
  \end{equation}
  where the $\theta_k$ are values drawn from the base distribution and the weights
  $\beta_k$ come from a stick-breaking process parameterized by the concentration 
  parameter $\gamma$ (equivalently referred to as GEM($\gamma$)).~\cite{halmos_random_1944} In shorthand, this
  specification can be expressed as $G_0 \sim DP(\gamma, H)$. The concentration 
  parameter, $\gamma$, expresses one's confidence in the base distribution $H$.
  We use a uniform base distribution for $H$. When $\gamma\to 0$, the first 
  weight of $G_0$, $\beta_1$, approaches unity and for $\gamma\to\infty$, the weights
  become uniform and $G_0$ closely resembles $H$. Each row, $G_j$, of the transition 
  matrix is produced by drawing from a DP specified using the $\beta$ vector as a 
  discrete base distribution and a separate concentration parameter, $\alpha$.
  \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
  \end{equation}
  This hierarchical specification ensures that the transition probabilities in 
  each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
  Once the model has converged only a finite number of states will have significant
  sampling.

@@

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a Python implementation of this process which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  %MRS2: give location for the code here where you introduce it.
  %BJC3: I emailed Emily Fox and she just asked me to use this citation. I'm happy to add a 
  %link to where she stores the code. Would that be best in the form of a reference?
  Therefore, we looked for convergence of the parameters in Equation~\ref{eqn:var}.
  We detected equilibration of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

  % BJC3: Cut this paragraph
%  We carefully applied the IHMM algorithm in a way which takes advantage of the
%  system's cylindrical symmetry. While there are a number of viable ways in which
%  one could choose to analyze these time series, we chose to use a multi-step 
%  procedure that we believe adequately distinguishes the diversity of dynamical 
%  behavior exhibited by the solutes. The final parameterization of the model is 
%  in cylindrical coordinates since each pore is a cylinder where we expect 
%  solutes to exhibit radially symmetric dynamics. We present a step-by-step 
%  demonstration of the procedure that follows in 
%  Section~\ref{S-section:ihmm_procedure} of the Supporting Information.

  We start by applying the IHMM algorithm to the 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center. We tracked the solute's
  motion along the pore axis with the center-of-mass $z$ coordinate. Using the nearest
  pore center as the origin, we represented the radial distance of each solute's 
  center-of-mass from the pore center in 2 dimensions, $x$ and $y$. By working in 
  Cartesian coordinates, we avoid mathematical complexity introduced by cylindrical 
  coordinates while estimating the state sequence.
  %BJC3: maybe could say something about multivariate guassian noise with VAR in cylindrical doesnt work with current code.

  We parameterized each of the 24 solute trajectories independently.
  Although the IHMM is capable of identifying an infinite number of states, 
  a Dirichlet process tends to exhibit a ``rich get richer" effect, favoring
  a fewer number of states.~\cite{dreyer_discovering_2011} By applying the algorithm to each trajectory 
  independently, we reduce the possibility of lumping together multiple 
  similar states which we would prefer to stay separated before clustering.
  Note that the states identified by the IHMM are heavily influenced by 
  the Gaussian prior placed on $\mathbf{c}$ in Equation~\ref{eqn:var}. 
  In Section~\ref{S-section:prior_guesses} of the Supporting Information we 
  outline our method for choosing reliable prior parameters.

@@

  % BJC3: This paragraph no longer necessary
%  We reparameterized the time series, preserving the state sequence, in terms
%  of the radial and axial coordinates ($r$, $z$) due to the system's radial 
%  symmetry. We converted the $x$ and $y$ center-of-mass coordinates to $r$. 
%  We fixed the state sequence from the 3D parameterization and applied the 
%  inference component of the IHMM procedure to the cylindrical trajectories 
%  in order to estimate the VAR(1) parameters in terms of $r$ and $z$. Since
%  we fixed the state sequence, and we are using conjugate priors, the VAR 
%  parameters converge very quickly (see Figure~\ref{S-fig:fixed_state_convergence}).
%  Therefore, we only ran the inference procedure for 100 iterations. 
%  % MRS: hmm. Here the new covariance parameters is redone in r - does there need to be any transformation applied? 
%  % MRS: though might be OK, since x^2 + y^2 = r^2 so maybe covariance are additive if x and y uncorrelated (which they should be)
%  % MRS2: As I mentioned before, the issue is if the likelihood for the covariance matrix is properly defined for radial diffusion.  We can check over the equations.  It likely wouldn't change the qualitative results much, but should ne checked.  Would not affect the clustering.
%  We defined the final parameters of each state as the mean of the parameters 
%  from each iteration recorded after the equilibration time point. We found the 
%  equilibration time point of each component of the $\mathbf{c}$ vector as well as 
%  $A$ and $\Sigma$ matrices, and then used the longest equilibration time of all 
%  dimensions as the equilibration time point.

  \subsubsection*{Predicting Mean Squared Displacement}\label{method:realizations}

  %BJC3: expanded
  \textit{Generating Stochastic IHMM Trajectory Realizations}: For each independent 
  MD trajectory, we generated stochastic trajectory realizations using the finalized 
  parameters of the associated IHMM. We drew state sequences with transition probabilities
  given by $T$. While in a given state, we simulated motion according to the VAR(1) 
  parameterization of that state. After each state transition, we set the unconditional
  mean of the new state to the particle's position immediately before the state 
  transition occurred.

  %BJC3: this is new, not sure of clarity
  \textit{Calculating MSD}: The MSD reported for each solute measured by our MD 
  simulations is a distribution created from the 24 independent MSD curves. We predicted 
  MSDs and their uncertainties with our stochastic models by bootstrapping the 
  distribution of stochastically generated MSDs. For each bootstrap trial, we generated
  24 independent trajectories using the 24 sets of IHMM parameters. We report the mean
  and standard error of the bootstrapped MSD distributions.

@@

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size. For each solute studied, we identified 200-325
  independent states, each with separate VAR(1) parameters. Many of these states
  exhibit very similar dynamical behavior except their mean levels in each dimension
  are different.

  We reduced the parameter space used agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage 
  criteria, which works to minimize the sum of the squared differences within all
  clusters.~\cite{ward_hierarchical_1963} We elected to choose the number of clusters
  rather than the distance threshold. For our data, non-parametric methods such as 
  Bayesian Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013}
  tend to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information), which could cause the clustered model to combine 
  distinct behavior into a non-existent hybrid state.
  %MRS: not clear what the consequence of the above is?
  %BJC3: added a little to the end

  %BJC3: reworked the following and split into paragraphs
  We clustered based on the eigenvalues of $A$ and $\Sigma$, the distance from the 
  pore center and the self-transition probabilities, $T_{ii}$, of each state. $A$ 
  and $\Sigma$ capture the correlation and size of fluctuations. $T_{ii}$ is directly
  related to the average time spent in each state. The distance from the closest pore 
  center is likely meaningful since the membrane is radially inhomogeneous from each 
  pore, transitioning from hydrophilic pores to hydrophobic tails. However, the 
  channels are isotropic along the pore axis so we do not in means in the $z$ dimension.
  We normalized each clustering dimension by its variance and shifted the mean to zero.

  To take advantage of the system's cylindrical symmetry, we converted the parameters that
  are directly related to solute position into cylindrical coordinates before clustering. We 
  converted the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ as $\lambda_x$ + $\lambda_y$
  and $\lambda_x^2$ + $\lambda_y^2$ respectively. %BJC3: possibly could use some justification of these definitions. 
  We obtained means in $r$, $\mu_r$, using the norm of the $x$ and $y$ means, 
  $\sqrt{\mu_x^2 + \mu_y^2}$. We squared $\mu_r$ for clustering in order to cast it in the
  same units as $A$ and $\Sigma$.

  To cluster on $T_{ii}$, we first cast it in terms of the expected dwell times. The expected
  value of the number of sequential self-transitions is simply $\frac{1}{1 - T_{ii}}$. This 
  relationship correctly implies that dwell times approach infinity as $T_{ii}$ approaches
  1. Therefore we found the most success with agglomerative clustering by linearizing this
  relationship and clustering on $-\log(1 - T_{ii})$ (see Section~\ref{S-section:selfT_cluster}
  of the Supporting Information for a graphical illustration).

@@

@@

  %BJC3: modified
  We remapped and fixed the state sequence based on the cluster assignments and
  parameterized the clusters using the inference component of the IHMM algorithm. 
  First, we modified the ($x$, $y$, $z$) solute trajectories so that they had a 
  mean of zero, leaving only the fluctuations. First, we zeroed the $y$ component
  of the trajectories by rotating the coordinates about the pore axis such that 
  the mean vector of each separate segment was aligned with the $x$ axis. Next, 
  we subtracted the means from each separate segment in the $x$ and $z$ dimensions.
  We used the IHMM algorithm on this modified trajectory to infer the clustered 
  state parameters by fixing the clustered state sequence.

  We obtained $\mathbf{c}$ vectors of the clustered states by averaging 
  each value of the rotated $\mathbf{c}$ vectors assigned to the same cluster. Note
  that we only care about the $x$ and $y$ component of $\mathbf{c}$ because solute 
  trajectories are not bound in the $z$ direction.

@@

  %We developed a way to 
  We measure the local density of the membrane at arbitrary
  points in the unit cell as follows. We histogrammed the positions in three dimensions
  Since our system is in a monoclinic unit cell, we periodically replicated
  the system in the $\pm x$, $y$ and $z$ directions and then chose the bounds on the
  histogram in order to create a rectangular box encompassing the unit cell with
  a 1 nm buffer between the histogram and unit cell boundaries.~\cite{van_der_walt_numpy_2011} We then used a regular
  grid interpolator in order to allow interpolation at arbitrary points within the grid.~\cite{virtanen_scipy_2020}

@@

  \begin{figure}
  \centering
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_MET.pdf}
  \caption{methanol}\label{fig:unclustered_msd_MET}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_GCL.pdf}
  \caption{ethylene glycol}\label{fig:unclustered_msd_GCL}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_URE.pdf}
  \caption{urea}\label{fig:unclustered_msd_URE}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_ACH.pdf}
  \caption{acetic acid}\label{fig:unclustered_msd_ACH}
  \end{subfigure}
  \caption{The MD MSDs (black) of methanol and urea are under-predicted by 
  the IHMM (orange), while the predictions match within uncertainty for ethylene
  glycol and urea.}\label{fig:unclustered_msds}
  %MRS2: to save space, could consider the graphs in 1x4 across the top, with a single legend (since all have the same), and slightly thicker lines  %There are only 2 lines per plot, so it's not too noisy to be that size
  %BJC3: good point
  \end{figure}

  %MRS2: I'd make this a separate subsection ``Deviations of the IMMM from molecular motion'' or something like that.
  % BJC3: would these subsubsection headers look better as just italicized and on 
  % the same line as where the next paragraph starts?
  \subsubsection*{Deviations of the IHMM from Molecular Motion}
